Formal Modelling of Predator Preferences using Molecular Gut-Content Analysis

### 2015-04-03 paper/main.tex
html vignette, starting to address reviewer's comments

\documentclass[12pt]{article}
\input{TexOpts.tex}

\begin{document}
\subsection*{Title:}  Formal Modelling of Predator Preferences using Molecular Gut-Content Analysis

\subsection*{Author Information:}
Edward A. Roualdes$^{1,2}$, Simon Bonner$^1$, Thomas D. Whitney$^{3,*}$, and James D. Harwood$^3$

\begin{tabular}{ll}
  1 & Department of Statistics\\
  & University of Kentucky\\
  & Rm.~311 Multidisciplinary Science Building\\
  & 725 Rose Street\\
  & Lexington, KY 40536-0082\\
  & USA\\
  \\
  2 & Email: edward.roualdes@uky.edu\\
  & Phone: 530-570-2674\\
  \\
  3 & Department of Entomology\\
  & University of Kentucky \\
  & Lexington, KY 40546\\
  & USA \\
  \\
  * & Current address: \\
  & Warnell School of Forestry and Natural Resources \\
  & University of Georgia \\
  & Athens, GA 30602\\
  & USA \\
  \\

\end{tabular}

\subsection*{Short Title:}
Modelling Predator Preferences

\subsection*{Word Count:}
This manuscript contains $3709$ words including figure captions and headers, but excluding title page and equations.

\subsection*{Keywords}
electivity; expectation-maximization; predator-prey interactions; generalist predators; food web analysis

\newpage
\section*{Summary}
\label{sec:abstract}

\begin{enumerate}
\item The literature on modelling a predator's prey selection describes many intuitive indices, few of which have both reasonable statistical justification and tractable asymptotic properties.
\item Here, we provide a simple model that meets both of these criteria, while extending previous work to include an array of data from multiple species and time points.
\item Further, we apply the expectation-maximisation algorithm to compute estimates if exact counts of the number of prey species eaten in a particular time period are not observed.
\item We conduct a simulation study to demonstrate the accuracy of our method, and illustrate the utility of the approach for field analysis of predation using a real data set, collected on wolf spiders using molecular gut-content analysis.
\end{enumerate}

\section{Introduction}
\label{sec:intro}

The indices most commonly used to describe a predator's food preferences, or selectivity, are relatively old \citep{Ivlev:1964,Jacobs:1974,Chesson:1978,Strauss:1979,Vanderploeg:1979,Chesson:1983}, and yet many applied papers continue to use them.  A quick search of papers published in $2014$ returns hundreds of publications that cite these fundamental papers, a few being \citet{Clements:2014,Hansen:2014,Hellstrom:2014,Lyngdoh:2014,Madduppa:2014}.  These indices, though intuitive, lack the statistical rigour of a full model, focus on a snapshot in time, and rarely allow more than one prey species to be considered \citep{Lechowicz:1982}.  We propose an intuitive statistical model to estimate and statistically test differences in a predators' prey preferences across an array of time points and between multiple prey species.

A comprehensive overview by \citet{Lechowicz:1982}, later summarised by \citet{Manly:1992}, details the benefits and faults of the most popular indices.  According to these reviews, a majority of the indices give comparable results, save Strauss's linear index $L$, despite the fact that most of the methods differ by range and linearity of response.  While Lechowicz recommends one index, $E^*$ by \citet{Vanderploeg:1979} as the ``single best'' \citep{Lechowicz:1982}, albeit imperfect, index, Manly et al.\ instead take the approach of excluding the subset of indices which do not ``estimate any biologically meaningful value'' \citep{Manly:1992}.  \citet{Lechowicz:1982} recommends the index $E^*$, an element of the \citet{Manly:1992} suggested indices, because the index value $0$ denotes random feeding, the index has a range restricted to $[-1,1]$ (though $E^*=1$ is nigh impossible), and the index is based on the predator's choice of prey as a function of both the availability of the prey as well as the number of available prey types (assumed known).  The downside to this index is its lack of reasonable statistical properties \citep{Lechowicz:1982}, thus making the computation of standard errors, and hypothesis testing difficult.  This is, in fact, a common fault amongst most of the indices.

To encourage more formal statistical inference, and simultaneously generalise predators' selectivity to animal resource selection, \citet{Manly:1992} proposed the use of generalised linear models (GLM).  The well established literature on GLMs allows for hypothesis testing to replace the indices, by estimating the proportion of eaten prey species, relative to the abundance of that prey species, while using environmental variables as predictors.  The model we present here, while restricted to predators' preferences, is a compromise between these two extremes, indices and GLMs.  Our model offers formal hypothesis testing and inference similar to the GLMs of \citet{Manly:1992}, but also provides meaningful single number summaries of the predator's dietary preferences.  To do this, we estimate the rate at which a predator consumes the prey of interest instead of estimating the proportion of consumed to available prey.  Outcomes of our model, are that we give up the somewhat arbitrary preference for an index to have the range $[-1,1]$, and random feeding, now denoted by $1$, is formally testable across time points and across prey species.

Our model enables formal hypothesis testing and statistical inference, while being general enough to perform statistical tests across multiple species and time points.  This provides researchers a more detailed analysis of the predator's feeding preferences.  Further, because our model is based on underlying Poisson distributions, members of the well studied exponential family, we are able to estimate the parameters of interest even when exact counts of each prey species eaten within any given time period are not observed.  Instead, we rely on the researcher being able to detect prey DNA within the predator's gut \citep{Schmidt:2014,Raso:2014,Madduppa:2014} and make a simple binary conclusion: this predator ate some of that prey species during this time period, or did not.

@@

By modelling $\lambda_{st}, \gamma_{st}$ we are able to test claims about a predator's eating preferences.  Formal statistical statements about the relative magnitudes of the parameters $\boldsymbol{\lambda} = (\lambda_{11}, \ldots, \lambda_{ST})^t$ and $\boldsymbol{\gamma} = (\gamma_{11}, \ldots, \gamma_{ST})^t$ offer insights to the relative rates at which predators eat particular prey species.  We consider five variations on the relative magnitude of $c_{st} = \lambda_{st}/\gamma_{st}$.  These five hypotheses each allow $c_{st}$ to vary by time, prey species, both, or neither.  Because the five hypotheses are nested, a natural testing order is suggested in Figure~\ref{fig:hier}.

@@

\begin{figure}
  \centering
  \begin{tikzpicture}[->,>=stealth',shorten >=1pt,auto,node distance=2.8cm, semithick]

    \node[state] (A)                    {$1$};
    \node[state] (B) [below of=A]       {$c$};
    \node[state] (C) [below left of=B]  {$c_s$};
    \node[state] (D) [below right of=B] {$c_t$};
    \node[state] (E) [below left of=D]  {$c_{st}$};

    \path       (B) edge node {} (A)
                (C) edge node {} (B)
                (D) edge node {} (B)
                (E) edge node {} (C)
                (E) edge node {} (D);
  \end{tikzpicture}

  \caption{This hierarchy of hypotheses suggests the order in which the discussed models should be tested.  One begins with the most complex models at the bottom and sequentially, following the arrows, tests simpler hypotheses using the formal test described in Section~\ref{sec:test} until a final model is established.}
  \label{fig:hier}
\end{figure}

@@

The model developed here allows for the determination of a predator's eating preferences by testing simultaneously across an array of multiple prey species and time points.  This is achieved via a simple, but statistically powerful, likelihood ratio test.  Further testing of the ratio of rates for which predators eat to encounter prey species allows researchers to make specific conclusions about predators' preferences.  For instance, rates across time can be estimated to make statements about seasonal effects on a predator's eating habits, while relative rates across species groups allows for statements about the relative preferences for different species.

When counts of predators' gut contents are not fully observed, and instead only a binary response indicating the existence of the prey species in the gut is observed, we are able to treat the counts as missing data.  By modelling all of the observed data, both the binary responses and the number of prey species caught, and the missing count data, we are able to use the EM algorithm to extract as much information from the data as possible.

@@

\section{Acknowledgements}
\label{sec:acknowledge}

The information reported in this paper (No.\ $15-08-008$) is part of a project of the Kentucky Agricultural Experiment Station and is published with the approval of the Director.  Support for this research was provided by the University of Kentucky Agricultural Experiment Station State Project KY008055 and the National Science Foundation Graduate Research Fellowship Program.

\section{Data Accessibility}
\label{sec:accessibility}

An \texttt{R} package, named \texttt{spiders}, is available on CRAN at \url{http://cran.r-project.org/web/packages/spiders/index.html} and fits all the methods discussed above.  \nocite{Wickham:2011a,Wickham:2011,Bischl:2014}

\bibliographystyle{plainnat}
\bibliography{refs}
\end{document}

>>>>

\documentclass[smallextended]{svjour3}
\input{TexOpts.tex}
\begin{document}
\title{Formal Modelling of Predator Preferences using Molecular Gut-Content Analysis}
\author{Edward A. Roualdes \and Simon Bonner \and Thomas D. Whitney \and James D. Harwood}
\institute{E.A. Roualdes \and S. Bonner
        \at Department of Statistics, University of Kentucky, Rm.~311 Multidisciplinary Science Building, 725 Rose Street, Lexington, KY 40536-0082, \\\email{edward.roualdes@uky.edu} \and
        T.D. Whitney 
        \at Department of Entomology, University of Kentucky, Lexington, KY 40546, USA \emph{Present address: Warnell School of Forestry and Natural Resources, University of Georgia, Athens, GA 30602, USA}
        \and
        J.D. Harwood
        \at Department of Entomology, University of Kentucky, Lexington, KY 40546, USA
}

% \subsection*{Keywords}
% 
\maketitle

\begin{abstract}
  The literature on modelling a predator's prey selection describes many intuitive indices, few of which have both reasonable statistical justification and tractable asymptotic properties.  Here, we provide a simple model that meets both of these criteria, while extending previous work to include an array of data from multiple species and time points.  Further, we apply the expectation-maximisation algorithm to compute estimates if exact counts of the number of prey species eaten in a particular time period are not observed.  We conduct a simulation study to demonstrate the accuracy of our method, and illustrate the utility of the approach for field analysis of predation using a real data set, collected on wolf spiders using molecular gut-content analysis.
  \keywords{electivity \and expectation-maximization \and predator-prey interactions \and generalist predators \and food web analysis}
\end{abstract}

% tweet: Formal modelling of a predator's prey selection across multiple species and time points using molecular gut-content analysis.

\section{Introduction}
\label{intro}

The indices most commonly used to describe a predator's food preferences, or selectivity, are relatively old \citep{Ivlev:1964,Jacobs:1974,Chesson:1978,Strauss:1979,Vanderploeg:1979,Chesson:1983}, and yet many applied papers continue to use them.  A quick search of papers published in $2014$ returns hundreds of publications that cite these fundamental papers, a few being \citet{Clements:2014,Hansen:2014,Hellstrom:2014,Lyngdoh:2014,Madduppa:2014}.  These indices, though intuitive, lack the statistical rigour of a full model, focus on a snapshot in time, and rarely allow more than one prey species to be considered \citep{Lechowicz:1982}.  Other authors \citep{Davey:2013,King:2010} are using the Monte Carlo based method of \citet{Agusti:2003}, but this also neglects to take into account multiple species across multiple time points.  We propose an intuitive statistical model to estimate and statistically test differences in a predators' prey preferences across an array of time points and between multiple prey species.

A comprehensive overview by \citet{Lechowicz:1982}, later summarised by \citet{Manly:2002}, details some of the benefits and faults of the most popular indices.  According to these reviews, a majority of the indices give comparable results, save Strauss's linear index $L$, despite the fact that most of the methods differ by range and linearity of response.  While Lechowicz recommends one index, $E^*$ by \citet{Vanderploeg:1979} as the ``single best'' \citep{Lechowicz:1982}, albeit imperfect, index, Manly et al.\ instead take the approach of excluding the subset of indices which do not ``estimate any biologically meaningful value'' \citep{Manly:2002}.  \citet{Lechowicz:1982} recommends the index $E^*$, an element of the \citet{Manly:2002} suggested indices, because the index value $0$ denotes random feeding, the index has a range restricted to $[-1,1]$ (though $E^*=1$ is nigh impossible), and the index is based on the predator's choice of prey as a function of both the availability of the prey as well as the number of available prey types (assumed known).  The downside to this index is its lack of reasonable statistical properties \citep{Lechowicz:1982}, thus making the computation of standard errors and thus hypothesis testing difficult.  This is, in fact, a common fault amongst most of the indices.

\citet{Manly:2002} recognized the need for more formal statistical inference and proposed the use of generalised linear models (GLM).  The well established literature on GLMs allows for formal hypothesis testing.  Much of this work however focues on general resource selection and doesn't directly address the problems of food selection.  Our model, while focusing on predators' food selection not general resource selection, offers formal hypothesis testing and inference similar to the GLMs of \citet{Manly:2002}, but also provides meaningful single number summaries of the predator's dietary preferences.  Thus, we are proposing here a model that 1) maintains the intuitiveness of the indices summarised by \citet{Lechowicz:1982, Manly:2002}, 2) offers the formal statistical justification of GLMs, 3) models multiple prey species across multiple time points, and 4) estimates parameters of interest even when exact counts of prey species eaten are not fully observed.

% To do this, we estimate the rate at which a predator consumes the prey of interest instead of estimating the proportion of consumed to available prey.  Outcomes of our model, are that we give up the somewhat arbitrary preference for an index to have the range $[-1,1]$, and random feeding, now denoted by $1$, is formally testable across time points and across prey species.

Our model is based on two Poisson distributions with parameters describing the rates at which the predator consumes prey species and the rate at which the predator encounters prey species.  Because the Poisson distirbution is a member of the well studied exponential family, we are able to estimate the parameters of interest even when exact counts of each prey species eaten within any given time period are not observed.  Instead, we rely on the researcher being able to detect prey DNA within the predator's gut \citep{Schmidt:2014,Raso:2014,Madduppa:2014} and make a simple binary conclusion: this predator ate some of that prey species during this time period, or did not.

@@

By modelling $\lambda_{st}, \gamma_{st}$ we are able to test claims about a predator's eating preferences.  Formal statistical statements about the relative magnitudes of the parameters $\boldsymbol{\lambda} = (\lambda_{11}, \ldots, \lambda_{ST})^t$ and $\boldsymbol{\gamma} = (\gamma_{11}, \ldots, \gamma_{ST})^t$ offer insights to the relative rates at which predators eat particular prey species.  We consider five variations on the relative magnitude of $c_{st} = \lambda_{st}/\gamma_{st}$.  These five hypotheses each allow $c_{st}$ to vary by time, prey species, both, or neither.

@@

\begin{figure}
  \centering
  \includegraphics{hyps.pdf}
  \caption{This hierarchy of hypotheses suggests the order in which the discussed models should be tested.  One begins with the most complex models at the bottom and sequentially, following the arrows, tests simpler hypotheses using the formal test described in Section~\ref{sec:test} until a final model is established.}
  \label{fig:hier}
\end{figure}

Because the five hypotheses are nested, a natural testing order is suggested in Figure~\ref{fig:hier}.  For instance, suppose interest lies in the predator's dietary preferences with respect to prey species $s_1,s_2$ across $3$ time periods.  If it is believed that hypothesis $4.$ best describes the predator's eating, i.e.\ that the prey species are equally preferred, such that $s_1,s_2$ can be thought of as one prey, but that this preference changes across time, then

\begin{align*}
  H_0&: \lambda_t = c_t \gamma_t, \forall t \\  
  H_1&: \lambda_{st} = c_{st}\gamma_{st}
\end{align*}

would test, statistically, if the data provide evidence for such a belief beyond the model that treats each prey species $s_1$ and $s_2$ individually across the $3$ time points.  Similarly,

\begin{align*}
  H_0&: \lambda_s = c_s \gamma_s, \forall s \\  
  H_1&: \lambda_{st} = c_{st}\gamma_{st}
\end{align*}

would test if $P$'s eating varies by prey species, but not by time, against the most parameter rich model $c_{st}$.  The above two tests should be performed before any of the simpler methods are tested.

@@

All the methods mentioned here, whether they be others' methods or our model, are susceptible to a number of issues that must be handled as best as possible by the investigator.  The most obvious are sampling issues of prey and predator species of interest.  It is almost impossible to know how well one's data represents the habitat and the availability of the prey of interest.  Variables not recorded when collecting data could have influenced the relative abundance of species that were caught.  And yet this is just the beginning.  Other practical challenges of analysing the data once in the lab provide immense opportunity for data to be comprimised.  We therefore, in this paper, focus solely on the methodologies themselves, assuming the best of possible scenarios brought your data to our model.

The earliest attempts to estimate predators' dietary preferences, summarised by \citet{Lechowicz:1982} and \citet{Manly:2002},  focused on producing one number summaries that rarely considered more than one prey species of interest, did not take into account changes across time, and had minimal statistical justification.  Instead, the indices developed were justified by arguing in favor of each index's unique, and claimed ``optimal,'' properties.  We feel these indices missed the point; an index is only as good as it is understood to estimate and model our world of observable data.

A more recent and notable method to esimate predator's preferences uses a Monte Carlo, or bootstrap to be more accurate, approach \citep{Agusti:2003}.  This method deserves credit for its lack of distributional assumptions, and yet it still misses the benefits of formally modelling multiple prey species and multiple time points.  Further, it should be noted that this Monte Carlo approach though creative and widely applicable, easily confuses a key statistical idea: re-sampling does not in any way increase one's sample size or the information contained there within.

The model presented here is a formal statistical model, which enables hypothesis testing of multiple prey species across an array of time points.  When appropriately fit to data our methods offer greater statistical power than any index or distribution-free Monte Carlo method could.  We were able to justify our method with robust simulations that provide insight into the statistical properties of our model.  Moreover, we offer the means to estimate parameters of interest even when the predator's gut contents are not fully observed.

Throughout this paper we assume the observed data were generated from Poisson distributions.  This enabled us to use the expectation-maximisation algorithm to fit our model when only incomplete data is observed.  Of course other distributional assumptions could be made in the case of fully observed count data, but it is not clear if this is true in the case of the unobserved count data.  For instance, consider simply replacing the Poisson distribution with a distribution with more than one parameter.  Since, in the case of non-count data, only $0$s and $1$s are observed, an underlying distribution with more than one parameter would be over-identified.  The only other option is to specify a distribution that better fits the observed data than the Poisson distribution, but here one needs to be careful about underyling processes that are being modelled.

Need words about users of these methods.

@@

\begin{acknowledgements}
The information reported in this paper (No.\ $15-08-008$) is part of a project of the Kentucky Agricultural Experiment Station and is published with the approval of the Director.  Support for this research was provided by the University of Kentucky Agricultural Experiment Station State Project KY008055 and the National Science Foundation Graduate Research Fellowship Program.  
\end{acknowledgements}
% \section{Data Accessibility}
% \label{sec:accessibility}

% An \texttt{R} package, named \texttt{spiders}, is available on CRAN at \url{http://cran.r-project.org/web/packages/spiders/index.html} and fits all the methods discussed above.  \nocite{Wickham:2011a,Wickham:2011,Bischl:2014}

\bibliographystyle{spbasic}
\bibliography{refs}
\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:
